Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-23 manuscript/main.Rmd
revision

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30m GFC resolution. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. Where the datasets overlap we provide summary statistics of deforestation, reforestation and afforestation.

@@

We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200m, with the 50m side of the plot along the forest edge and continuing 200m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

@@

Scaling our classifier to the whole historical orthomosaic we detected **16200 ha** (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two **acquisition periods**. In total, **9900 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate **8800 ha (Table 3, Figure 6 - light green)**.

@@

A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05). Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13200 ha**.

@@

**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes. The methodology is analoguous to the use of sparse labelling as used by [@buscombe2018] and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis. More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015]. The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic. More so, our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time. We acknowledge that our approach will speed up labelling, but is limited by the available homogeneous areas to sample from and the number of classes. Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**

@@

Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**16200 ha**). Considerable regrowth has occured since the aerial survey was executed (**~9900 ha**), and socio-economic instability prevented further large scale forest exploitation. In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas. The majority of this forest recovery takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest. Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300m from the edge [@gascon2000]. Our analysis of edge effects on AGC has shown that the influence is negligible 200m away from the edge. @phillips2006 have shown similar weak responses to edge effects in the Amazon forest. Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200m we did not include any estimates of carbon loss or gain within these zones. However, it must be stated that the influence of edges throughout the landscape was not marginal. as these areas would account for **13200 ha**. Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC. Similarly, uncertainties in how to explicitly correct for plantations in the landscape present a further challenge. Thus although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting. However, our findings do not indicate that deforestation in Congo basin is declining, on the contrary.

@@

Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change especially in the pre-satellite era is an imperative. Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene. Efforts to quantify canopy texture and their link to carbon storage had limited to no success.We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks. We also show a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. As such, our analysis provides insights into the state of rarely studied tropical forests and the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin providing a useful historical context to land-cover and land-use change spatial patterns for interpreting past and ongoing field research in the area.

>>>>

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution **using a nearest neighbour approach**. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. We provide summary statistics of historical and contemporary forest loss and regrowth.

deforestation (persistent forest loss after 1958), reforestation (regrowth after loss in 1958).

@@

We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

@@

Scaling our classifier to the whole historical orthomosaic we detected **~16200 ha** (or ~20% of the scene) of disturbed forests. A large fraction of the disturbed area was restored in the period between the two **acquisition periods**. In total, **9918 ha**, or little over half of the affected forest was restored (Figure 5C-D, dark blue). Recent deforested areas, as registered through satellite remote sensing, approximate **8776 ha (Table 3, Figure 6 - light green)**.

@@

A comparison of forest edge plots with mixed forest plots showed no signficant difference in AGC, or other reported values such as species richness, basal area or stem density (Mann Whitney U test, p < 0.05). Edge influence did not extend beyond 200 m from a forest edge, but still represented an area of **13151 ha**.

@@

**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes. The methodology is analoguous to the use of sparse labelling as used by @buscombe2018 and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis. More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015]. The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic. Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time. We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes. Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**

@@

Our analysis shows that the majority of deforestation around Yangambi happened toward the late 1950's (**~16200 ha**). Considerable regrowth has occured since the aerial survey was executed (**~9918 ha**), and socio-economic instability prevented further large scale forest exploitation. In particular, many plantations have reached maturity and forest has re-established in previously cleared or disturbed areas. The majority of this forest recovery takes the form of isolated patches of forest but is offset by further deforestation of previously untouched forest. Generally, the function and structure of forests can be influenced by forest edges that are located up to 1km away however most effects are pronounced within the first 300m from the edge [@gascon2000]. Our analysis of edge effects on AGC has shown that the influence is negligible 200m away from the edge. @phillips2006 have shown similar weak responses to edge effects in the Amazon forest. Due to a lack of data on the extent (depth) of edge effects and their influence on AGC beyond 200m we did not include any estimates of carbon loss or gain within these zones. However, it must be stated that the influence of edges throughout the landscape was not marginal. as these areas would account for **13151 ha**. Thus edges could have a substantial negative [@brinck2017] or positive [@reinmann2017] influence on AGC. Similarly, uncertainties in how to explicitly correct for plantations in the landscape present a further challenge. Thus although our estimates are only indicative they do underscore the important influence of landscape structure when carbon accounting. However, our findings do not indicate that deforestation in Congo basin is declining, on the contrary.

@@

Given the impact of tropical forest disturbances on atmospheric CO~2~ emissions, biodiversity and ecosystem productivity accurate long term reporting on land-cover and land-use change in the pre-satellite era is an imperative. Our analysis of historical aerial survey images of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. We have shown that historical aerial survey data can be used to quantify land-use and land-cover change using a highly automated workflow to quantify the structure and extent of forest cover change that can help assess the impact of fragmentation on above ground carbon stocks. We also show a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportuntiy for substantial regrowth. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. As such, our analysis provides insights into the state of rarely studied tropical forests and the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin providing a useful historical context to land-cover and land-use change spatial patterns for interpreting past and ongoing field research in the area.

### 2020-01-23 manuscript/main.tex
revision

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 (GFC, tile
10N-020E) map data \citep{hansen2013}. Using the \textbf{GFC} data we
calculated the latest state of the forest with respect to the conditions
at the start of 1958, 60 years earlier. In our analysis we only included
\textbf{GFC} pixels which recorded no \textbf{forest} loss throughout
the whole 2000 - 2018 period. As the resolution of the historical forest
classification exceeds that of the GFC map we downsampled our historical
forest cover data to 30m GFC resolution. We masked out all water bodies
using the Global Forest Change survey data mask layer, and limited the
analysis to the right bank of the Congo river. Where the datasets
overlap we provide summary statistics of deforestation, reforestation
and afforestation.

@@

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots edge plots were set back 200m from forest
edges and were 50x200m, with the 50m side of the plot along the forest
edge and continuing 200m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

@@

Scaling our classifier to the whole historical orthomosaic we detected
\textbf{16200 ha} (or \textasciitilde{}20\% of the scene) of disturbed
forests. A large fraction of the disturbed area was restored in the
period between the two \textbf{acquisition periods}. In total,
\textbf{9900 ha}, or little over half of the affected forest was
restored (Figure 5C-D, dark blue). Recent deforested areas, as
registered through satellite remote sensing, approximate \textbf{8800 ha
(Table 3, Figure 6 - light green)}.

@@

A comparison of forest edge plots with mixed forest plots showed no
signficant difference in AGC, or other reported values such as species
richness, basal area or stem density (Mann Whitney U test, p \textless{}
0.05). Edge influence did not extend beyond 200 m from a forest edge,
but still represented an area of \textbf{13200 ha}.

@@

\textbf{Our approach uses broadly defined homogeneous polygons to
construct a balanced dataset of synthethic landscapes. The methodology
is analoguous to the use of sparse labelling as used by
\citep{buscombe2018} and contrasts with the standard methodologies which
generally extract pixel (windows) \citep{song2015} or delineate land
cover classes \citep{nita2018} to drive a classifier or analysis. More
so, the use of heavy image augmentation during model training sidesteps
texture representation issues which affect classification of image
scenes with inconsistent illumination or sharpness \citep{hudak1998} or
ad-hoc feature engineering \citep{song2015}. The use of synthetic
landscapes allowed us to account for most, but not all, of the
variability within our orthomosaic. More so, our analysis has shown that
despite being trained on historical data our model could map
contemporary forest cover in remote sensing data with similar spatial
and spectral characteristics (Figure 7), suggesting that the classifier
consistently across works both space and time. We acknowledge that our
approach will speed up labelling, but is limited by the available
homogeneous areas to sample from and the number of classes. Yet, the
latter should not be a constrained as previous research efforts have
focussed on simple forest loss maps \citep{hansen2013}.}

@@

Our analysis shows that the majority of deforestation around Yangambi
happened toward the late 1950's (\textbf{16200 ha}). Considerable
regrowth has occured since the aerial survey was executed
(\textbf{\textasciitilde{}9900 ha}), and socio-economic instability
prevented further large scale forest exploitation. In particular, many
plantations have reached maturity and forest has re-established in
previously cleared or disturbed areas. The majority of this forest
recovery takes the form of isolated patches of forest but is offset by
further deforestation of previously untouched forest. Generally, the
function and structure of forests can be influenced by forest edges that
are located up to 1km away however most effects are pronounced within
the first 300m from the edge \citep{gascon2000}. Our analysis of edge
effects on AGC has shown that the influence is negligible 200m away from
the edge. \citet{phillips2006} have shown similar weak responses to edge
effects in the Amazon forest. Due to a lack of data on the extent
(depth) of edge effects and their influence on AGC beyond 200m we did
not include any estimates of carbon loss or gain within these zones.
However, it must be stated that the influence of edges throughout the
landscape was not marginal. as these areas would account for
\textbf{13200 ha}. Thus edges could have a substantial negative
\citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
Similarly, uncertainties in how to explicitly correct for plantations in
the landscape present a further challenge. Thus although our estimates
are only indicative they do underscore the important influence of
landscape structure when carbon accounting. However, our findings do not
indicate that deforestation in Congo basin is declining, on the
contrary.

@@

Given the impact of tropical forest disturbances on atmospheric
CO\textsubscript{2} emissions, biodiversity and ecosystem productivity
accurate long term reporting on land-cover and land-use change
especially in the pre-satellite era is an imperative. Our analysis of
historical aerial survey images of the Central Congo Basin provides a
window into the state of the forest at the start of the anthropocene.
Efforts to quantify canopy texture and their link to carbon storage had
limited to no success.We have shown that historical aerial survey data
can be used to quantify land-use and land-cover change using a highly
automated workflow to quantify the structure and extent of forest cover
change that can help assess the impact of fragmentation on above ground
carbon stocks. We also show a shift from previously highly structured
industrial deforestation of large areas \textbf{for plantation
purposes}, to discrete smallholder clearing for farming, increasing
landscape fragmentation but also opportuntiy for substantial regrowth.
Efforts to quantify canopy texture and their link to carbon storage had
limited to no success. As such, our analysis provides insights into the
state of rarely studied tropical forests and the rate at which
deforestation and reforestation has taken place over a multi-decadal
scale in the central Congo basin providing a useful historical context
to land-cover and land-use change spatial patterns for interpreting past
and ongoing field research in the area.

>>>>

To map long term land-use and land-cover change in the Yangambi region
we used the contemporary Global Forest Change version 1.6 (GFC, tile
10N-020E) map data \citep{hansen2013}. Using the \textbf{GFC} data we
calculated the latest state of the forest with respect to the conditions
at the start of 1958, 60 years earlier. In our analysis we only included
\textbf{GFC} pixels which recorded no \textbf{forest} loss throughout
the whole 2000 - 2018 period. As the resolution of the historical forest
classification exceeds that of the GFC map we downsampled our historical
forest cover data to 30 m GFC resolution \textbf{using a nearest
neighbour approach}. We masked out all water bodies using the Global
Forest Change survey data mask layer, and limited the analysis to the
right bank of the Congo river. We provide summary statistics of
historical and contemporary forest loss and regrowth.

deforestation (persistent forest loss after 1958), reforestation
(regrowth after loss in 1958).

@@

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots edge plots were set back 200m from forest
edges and were 50x200 m, with the 50 m side of the plot along the forest
edge and continuing 200 m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200 m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

@@

Scaling our classifier to the whole historical orthomosaic we detected
\textbf{\textasciitilde{}16200 ha} (or \textasciitilde{}20\% of the
scene) of disturbed forests. A large fraction of the disturbed area was
restored in the period between the two \textbf{acquisition periods}. In
total, \textbf{9918 ha}, or little over half of the affected forest was
restored (Figure 5C-D, dark blue). Recent deforested areas, as
registered through satellite remote sensing, approximate \textbf{8776 ha
(Table 3, Figure 6 - light green)}.

@@

A comparison of forest edge plots with mixed forest plots showed no
signficant difference in AGC, or other reported values such as species
richness, basal area or stem density (Mann Whitney U test, p \textless{}
0.05). Edge influence did not extend beyond 200 m from a forest edge,
but still represented an area of \textbf{13151 ha}.

@@

\textbf{Our approach uses broadly defined homogeneous polygons to
construct a balanced dataset of synthethic landscapes. The methodology
is analoguous to the use of sparse labelling as used by
\citet{buscombe2018} and contrasts with the standard methodologies which
generally extract pixel (windows) \citep{song2015} or delineate land
cover classes \citep{nita2018} to drive a classifier or analysis. More
so, the use of heavy image augmentation during model training sidesteps
texture representation issues which affect classification of image
scenes with inconsistent illumination or sharpness \citep{hudak1998} or
ad-hoc feature engineering \citep{song2015}. The use of synthetic
landscapes allowed us to account for most, but not all, of the
variability within our orthomosaic. Our analysis has shown that despite
being trained on historical data our model could map contemporary forest
cover in remote sensing data with similar spatial and spectral
characteristics (Figure 7), suggesting that the classifier consistently
across works both space and time. We acknowledge that the use of
synthetic landscapes is limited by the available homogeneous areas to
sample from and the number of classes. Yet, the latter should not be a
constrained as previous research efforts have focussed on simple forest
loss maps \citep{hansen2013}.}

@@

Our analysis shows that the majority of deforestation around Yangambi
happened toward the late 1950's (\textbf{\textasciitilde{}16200 ha}).
Considerable regrowth has occured since the aerial survey was executed
(\textbf{\textasciitilde{}9918 ha}), and socio-economic instability
prevented further large scale forest exploitation. In particular, many
plantations have reached maturity and forest has re-established in
previously cleared or disturbed areas. The majority of this forest
recovery takes the form of isolated patches of forest but is offset by
further deforestation of previously untouched forest. Generally, the
function and structure of forests can be influenced by forest edges that
are located up to 1km away however most effects are pronounced within
the first 300m from the edge \citep{gascon2000}. Our analysis of edge
effects on AGC has shown that the influence is negligible 200m away from
the edge. \citet{phillips2006} have shown similar weak responses to edge
effects in the Amazon forest. Due to a lack of data on the extent
(depth) of edge effects and their influence on AGC beyond 200m we did
not include any estimates of carbon loss or gain within these zones.
However, it must be stated that the influence of edges throughout the
landscape was not marginal. as these areas would account for
\textbf{13151 ha}. Thus edges could have a substantial negative
\citep{brinck2017} or positive \citep{reinmann2017} influence on AGC.
Similarly, uncertainties in how to explicitly correct for plantations in
the landscape present a further challenge. Thus although our estimates
are only indicative they do underscore the important influence of
landscape structure when carbon accounting. However, our findings do not
indicate that deforestation in Congo basin is declining, on the
contrary.

@@

Given the impact of tropical forest disturbances on atmospheric
CO\textsubscript{2} emissions, biodiversity and ecosystem productivity
accurate long term reporting on land-cover and land-use change in the
pre-satellite era is an imperative. Our analysis of historical aerial
survey images of the Central Congo Basin provides a window into the
state of the forest at the start of the anthropocene. Efforts to
quantify canopy texture and their link to carbon storage had limited to
no success. We have shown that historical aerial survey data can be used
to quantify land-use and land-cover change using a highly automated
workflow to quantify the structure and extent of forest cover change
that can help assess the impact of fragmentation on above ground carbon
stocks. We also show a shift from previously highly structured
industrial deforestation of large areas \textbf{for plantation
purposes}, to discrete smallholder clearing for farming, increasing
landscape fragmentation but also opportuntiy for substantial regrowth.
Efforts to quantify canopy texture and their link to carbon storage had
limited to no success. As such, our analysis provides insights into the
state of rarely studied tropical forests and the rate at which
deforestation and reforestation has taken place over a multi-decadal
scale in the central Congo basin providing a useful historical context
to land-cover and land-use change spatial patterns for interpreting past
and ongoing field research in the area.
